Constructing a universe for the setoid model

### 2020-10-12 paper.tex
add Palmgren
Adapt paper and abstract to LNCS format

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2019}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{dsfont}
\usepackage{proof}
\usepackage{bbm}
%\usepackage[bw]{agda}

@@

@@

\author{Thorsten Altenkirch}{School of Computer Science, University of Nottingham, UK}{psztxa@nottingham.ac.uk}{https://orcid.org/0000-0002-6582-5025}{Supported by USAF grant FA9550-16-1-0029.}
\author{Simon Boulier}{Inria, Nantes, France}{simon.boulier@inria.fr}{}{Supported by ERC Starting Grant CoqHoTT 637339.}
\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no. ED\_18-1-2019-0030 and the Bolyai Fellowship of the Hungarian Academy of Sciences, Project no.\ BO/00659/19/3.}
\author{Christian Sattler}{Chalmers University of Technology, Gothenburg, Sweden}{sattler@chalmers.se}{}{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
\author{Filippo Sestini}{School of Computer Science, University of Nottingham, UK}{Filippo.Sestini@nottingham.ac.uk}{}{}

@@

\Copyright{Thorsten Altenkirch and Simon Boulier and Ambrus Kaposi and Christian Sattler and Filippo Sestini} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[500]{Theory of computation~Logic~Type theory} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{type theory, function extensionality, univalence, setoid model, induction-recursion, induction-induction} %TODO mandatory; please add comma-separated list of keywords

%\category{} %optional, e.g. invited paper

%\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{I want to thank \dots}%optional

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\maketitle

\begin{abstract}
The setoid model of type theory can be used to show the consistency of
intensional type theory with certain extensionality principles, like function
extensionality. In addition, Altenkirch's formulation of such model justifies a
universe of strict (definitionally-proof irrelevant) propositions supporting
propositional extensionality, a limited form of univalence according to which
logically equivalent propositions are equal. The appeal of this model
construction is that it takes place in a small, intensional, type theoretic
metatheory, therefore giving a method to boostrap extensionality.

Recently, Altenkirch's setoid model has been turned into a formal system, under
the name of Setoid Type Theory (SeTT). SeTT can be presented as an extension of
intensional Martin-Loef type theory with syntactic constructs that internalize
setoid reasoning. These give full access to the extensionality principles that
hold in the setoid model, while retaining the convenience of an intensional
theory.

As currently defined, SeTT is already a rich theory. However, its introspection
capabilities are currently lacking, since it only includes (univalent) universes
for propositions. We would like to extend SeTT with a universe of setoids
internalizing the notion of type beyond propositions. This amounts to extending
the setoid model with a semantics for such a universe.

In this paper, we present the construction of a (non-univalent) universe of
setoids within the setoid model. The universe is given in steps, first as an
inductive-recursive definition, which is then translated to an
inductive-inductive definition, and finally to an inductive family. Every step
is obtained from the previous by applying known reduction methods, although in a
novel way. Additionally, our construction shows the setoid model with the
additional encoding of the universe can still be defined in a relatively small
metatheory, that however must include a universe of strict propositions, indexed
W types, and a proof-irrelevant identity type with a strong transport rule. All
the mathematical developments described in the paper have been formalized in the
Agda proof-assistant.
\end{abstract}

@@

The setoid model was first described in \cite{hofmann} in order to add
extensionality principles to Type Theory such as function extensionality and
propositional extensionality. A strict variant of the setoid model was given in
\cite{setoid99} using a definitionally proof-irrelevant universe of
propositions. Recently, support for such a universe was added to the
proof-assistants Agda and Coq \cite{agda-prop}, allowing a full formalization of
Altenkirch's setoid model. Setoid Type Theory (SeTT) is a recently developed
formal system derived from this model construction \cite{mpc19}. Observational
Type Theory (OTT) \cite{alti:ott-conf, alti:ott-conf} is a syntax for the setoid
model differing from SeTT for the use of a different notion of heterogeneous
equality. Moreover, the consistency proof for OTT relies on Extensional Type
Theory, whereas for SeTT it is obtained via a syntactic translation.
%
XTT \cite{xtt} is a cubical variant of OTT where the equality type is defined
using an interval pretype. XTT's universes support universe induction, whereas
it is left open whether the construction presented here supports this principle.

@@

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{0}$, $\mathbf{1}$, $\Pi$, and $\Sigma$. The latter three can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
%
\begin{align*}
  \mathbf{1}_\mProp & :\equiv \| \mathbf{1} \| \\
  \Pi_\mProp\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_\mProp\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
\end{align*}
We assume that we have
$\mathbf{0}_\mProp:\mProp$ together with $\mathsf{exfalso}_\mProp :\{A:\mType \} \to \mathbf{0}_\mProp \to A$.

Finally, we will assume an identity type in the style of Martin-L\"of's
inductive identity type. The main difference is that our identity type is a
$\mProp$-valued relation. We have a transport combinator
\textsf{transp} from which \textsf{J} is derivable.
%
\begin{alignat*}{10}
  & \textsf{Id} && :\,\, && \{A : \mType \} \to A \to A \to \mProp \\
  & \textsf{refl} && : && \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
  & \textsf{transp} && : && \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1 \\
  & && && \textsf{transp}\ C\ \{x\}\ \{x\}\ e\ u \equiv u
\end{alignat*}

@@

The setoid model can be framed categorically as a category with families (CwF,
\cite{internal-tt}) with extra structure for the various type and term
formers. The core structure of a CwF can be given as the following signature:
\begin{align*}
  \Con &: \mType \\
  \Ty &: (\GG : \Con) \to \mType \\
  \Sub &: (\GG\ \GD : \Con) \to \mType \\
  \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
\end{align*}

@@

As an example, consider the following definition of a small universe with
bool and $\Pi$.

\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}

@@

In our case, we can encode $\U$ as an inductive type $\inU$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\inU$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
the type of codes, the definition of $\inU$ quite expectedly reflects the
inductive structure of codes.
%
\begin{alignat*}{10}
  & \textsf{data}\ \inU && : \mType \to \mType_1 \\
  & \quad \textsf{inBool} && : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi} && :  \inU\ A \to ((a : A) \to \inU\ (B\ a)) \to \inU\ ((a : A) \to (B\ a))
\end{alignat*}

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\
%
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
%
\begin{align*}
  & \El : \setoidU \to \mType \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
%
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
  & ... \\
  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\

@@

We will follow the method outlined in Section~\ref{ir-universes}. In
addition to $\inU$ for defining $\U$, we also introduce a family
$\inUU$ of binary relations between types in the universe, from which we
then define $\_\sim_{\setoidU}\_$.
%
\begin{align*}
  & \textsf{data}\ \inU : \mType \to \mType_1 \\
  & \textsf{data}\ \inUU : \{A\ A' : \mType\} \to \inU\ A \to \inU\ A' \to (A \to A' \to \mProp) \to \mType_1
\end{align*}

@@

As in the universe example in Section~\ref{ir-universes}, we now define
$\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
\begin{alignat*}{10}
  & \setoidU && : \mType_1 && \El : \setoidU \to \mType \\
  & \setoidU && :\equiv \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El :\equiv \pi_1
\end{alignat*}

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly isn't, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
%
\begin{align*}
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \mProp)\ (\inUU\ p\ p'\ R) \parallel
\end{align*}

We are now left to define the indexed equivalence relation on $\El$:
%
\begin{align*}
  & \_\vdash\_\sim_{\El}\_ : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
  & p \vdash a \sim_{\El} b :\equiv \ ?
\end{align*}

>>>>

%% \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{llncs}
\documentclass[autoref]{llncs}

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{dsfont}
\usepackage{proof}
\usepackage{bbm}
%\usepackage[bw]{agda}

@@

\begin{document}

@@

%% \author{Thorsten Altenkirch}{School of Computer Science, University of Nottingham, UK}{psztxa@nottingham.ac.uk}{https://orcid.org/0000-0002-6582-5025}{Supported by USAF grant FA9550-16-1-0029.}
%% \author{Simon Boulier}{Inria, Nantes, France}{simon.boulier@inria.fr}{}{Supported by ERC Starting Grant CoqHoTT 637339.}
%% \author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no. ED\_18-1-2019-0030 and the Bolyai Fellowship of the Hungarian Academy of Sciences, Project no.\ BO/00659/19/3.}
%% \author{Christian Sattler}{Chalmers University of Technology, Gothenburg, Sweden}{sattler@chalmers.se}{}{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
%% \author{Filippo Sestini}{School of Computer Science, University of Nottingham, UK}{Filippo.Sestini@nottingham.ac.uk}{}{}

\author{Thorsten Altenkirch\inst{1}
  \thanks{Supported by USAF grant FA9550-16-1-0029.}
  \orcidID{0000-0002-6582-5025} \and
  Simon Boulier\inst{2}\thanks{Supported by ERC Starting Grant CoqHoTT 637339.} \and
  Ambrus Kaposi\inst{3}\orcidID{0000-0001-9897-8936}
  \thanks{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no. ED\_18-1-2019-0030 and the Bolyai Fellowship of the Hungarian Academy of Sciences, Project no.\ BO/00659/19/3.}\and
  Christian Sattler\inst{4}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.} \and
  Filippo Sestini\inst{1}\orcidID{0000-0002-8701-5613}}

\institute{School of Computer Science, University of Nottingham, UK
  \email{\{psztxa,psxfs5\}@nottingham.ac.uk} \and
  Inria, Nantes, France\email{simon.boulier@inria.fr} \and
  E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary
  \email{akaposi@inf.elte.hu} \and
  Chalmers University of Technology, Gothenburg, Sweden
  \email{sattler@chalmers.se}
}

@@

\maketitle              % typeset the header of the contribution

%% \Copyright{Thorsten Altenkirch and Simon Boulier and Ambrus Kaposi and Christian Sattler and Filippo Sestini} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

%% \ccsdesc[500]{Theory of computation~Logic~Type theory} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\begin{abstract}
The setoid model is a model of intensional type theory that validates certain
extensionality principles, like function extensionality and propositional
extensionality, the latter being a limited form of univalence that equates
logically equivalent propositions. The appeal of this model construction is that
it can be constructed in a small, intensional, type theoretic metatheory,
therefore giving a method to boostrap extensionality.
%
The setoid model has been recently adapted into a formal system, namely Setoid
Type Theory (SeTT). SeTT is an extension of intensional Martin-L\"of type theory
with constructs that give full access to the extensionality principles that hold
in the setoid model.

Although already a rich theory as currently defined, SeTT currently lacks a way
to internalize the notion of type beyond propositions. This paper is a step
towards extending SeTT with a universe of setoids.
%
To this aim, we present the construction of a (non-univalent) universe of
setoids within the setoid model, first as an inductive-recursive definition,
which is then translated to an inductive-inductive definition and finally to an
inductive family. These translations from more powerful definition schemas to
simpler ones ensure that our construction can still be defined in a relatively
small metatheory, that however must include a proof-irrelevant identity type
with a strong transport rule. \keywords{type theory \and
  function extensionality \and univalence \and setoid model \and
  induction-recursion \and induction-induction}
\end{abstract}

@@

The setoid model was first described in \cite{hofmann} in order to add
extensionality principles to Type Theory such as function extensionality and
propositional extensionality. A strict variant of the setoid model was given in
\cite{setoid99} using a definitionally proof-irrelevant universe of
propositions. Recently, support for such a universe was added to the
proof-assistants Agda and Coq \cite{agda-prop}, allowing a full formalization of
Altenkirch's setoid model. Setoid Type Theory (SeTT) is a recently developed
formal system derived from this model construction \cite{mpc19}. Observational
Type Theory (OTT) \cite{alti:ott-conf, alti:ott-conf} is a syntax for the setoid
model differing from SeTT for the use of a different notion of heterogeneous
equality. Moreover, the consistency proof for OTT relies on Extensional Type
Theory, whereas for SeTT it is obtained via a syntactic translation.
%
XTT \cite{xtt} is a cubical variant of OTT where the equality type is defined
using an interval pretype. XTT's universes support universe induction, whereas
it is left open whether the construction presented here supports this principle.
%
Palmgren and Wilander \cite{2014arXiv1408.1364P} construct a setoid
universe using a translation into constructive set theory. In
contrast, we target a basic intensional type theory and hence also
give a computational semantics.

@@

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{0}$, $\mathbf{1}$, $\Pi$, and $\Sigma$. The latter three can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
\vspace{-0.2em}
\begin{align*}
  \mathbf{1}_\mProp & :\equiv \| \mathbf{1} \| \\
  \Pi_\mProp\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_\mProp\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
\end{align*}
We assume that we have
$\mathbf{0}_\mProp:\mProp$ together with $\mathsf{exfalso}_\mProp :\{A:\mType \} \to \mathbf{0}_\mProp \to A$.

Finally, we will assume an identity type in the style of Martin-L\"of's
inductive identity type. The main difference is that our identity type is a
$\mProp$-valued relation. We have a transport combinator
\textsf{transp} from which \textsf{J} is derivable.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \textsf{Id} && :\,\, && \{A : \mType \} \to A \to A \to \mProp \\
  & \textsf{refl} && : && \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
  & \textsf{transp} && : && \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1 \\
  & && && \textsf{transp}\ C\ \{x\}\ \{x\}\ e\ u \equiv u
\end{alignat*}

@@

The setoid model can be framed categorically as a category with families (CwF,
\cite{internal-tt}) with extra structure for the various type and term
formers. The core structure of a CwF can be given as the following signature:
\vspace{-0.2em}
\begin{align*}
  \Con &: \mType \\
  \Ty &: (\GG : \Con) \to \mType \\
  \Sub &: (\GG\ \GD : \Con) \to \mType \\
  \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
\end{align*}

@@

As an example, consider the following definition of a small universe with
bool and $\Pi$.
\vspace{-0.2em}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}

@@

In our case, we can encode $\U$ as an inductive type $\inU$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\inU$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
the type of codes, the definition of $\inU$ quite expectedly reflects the
inductive structure of codes.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \textsf{data}\ \inU && : \mType \to \mType_1 \\
  & \quad \textsf{inBool} && : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi} && :  \inU\ A \to ((a : A) \to \inU\ (B\ a)) \to \inU\ ((a : A) \to (B\ a))
\end{alignat*}

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
\vspace{-0.2em}
\begin{align*}
  & \El : \setoidU \to \mType \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
  & ... \\
  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\

@@

We will follow the method outlined in Section~\ref{ir-universes}. In
addition to $\inU$ for defining $\U$, we also introduce a family
$\inUU$ of binary relations between types in the universe, from which we
then define $\_\sim_{\setoidU}\_$.
\vspace{-0.2em}
\begin{align*}
  & \textsf{data}\ \inU : \mType \to \mType_1 \\
  & \textsf{data}\ \inUU : \{A\ A' : \mType\} \to \inU\ A \to \inU\ A' \to (A \to A' \to \mProp) \to \mType_1
\end{align*}

@@

As in the universe example in Section~\ref{ir-universes}, we now define
$\setoidU$ as a $\Sigma$ type, and $\El$ as the corresponding first projection.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \setoidU && : \mType_1 && \El : \setoidU \to \mType \\
  & \setoidU && :\equiv \Sigma\ (X : \mType)\ (\inU\ X) \hspace{5em} && \El :\equiv \pi_1
\end{alignat*}

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly isn't, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \mProp)\ (\inUU\ p\ p'\ R) \parallel
\end{align*}

We are now left to define the indexed equivalence relation on $\El$:
\vspace{-0.2em}
\begin{align*}
  & \_\vdash\_\sim_{\El}\_ : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
  & p \vdash a \sim_{\El} b :\equiv \ ?
\end{align*}
